Appendix S2. Evolutionary drivers of species and community dynamics

### 2024-11-30 manuscript/Appendix_S2_v02.Rmd
Updated mscript, all appendices, all figures.

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r setup,echo=FALSE,message=FALSE}
library(Hmsc)
library(ggplot2)
library(tidyr)
library(coda)
library(patchwork)
library(jtools)
library(bayesplot)
library(reshape2)
```
```{r,echo=FALSE,message=FALSE,warning=FALSE}
# Wrap long lines
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE,out.width="50%")
```

@@

We simulate population growth under particular parameter values, to generate population dynamics where species interactions, environment, and trait evolution all drive population dynamics.

```{r}
set.seed(42)
# Case 1: Weaker interactions, stronger environment
# Simulate initial species population growth with environment fluctuations
N1.0 <- 10
N2.0 <- 10
alpha.11 <- 0.01
alpha.22 <- 0.01
alpha.12 <- 0.005
alpha.21 <- 0.01
E.0 <- 0.8
x1.0 <- 0.1
x2.0 <- 0.8
P <- 1
w <- 2
Wmax <- 2
h2 <- 1
k <- (w + (1 - h2) * P)/(P + w)

# model function
disc_LV_evol <- function(N1.0,N2.0,alpha.11,alpha.22,alpha.12,alpha.21,E,x1.0,x2.0,P,w,Wmax,h2) {
  What <- Wmax*sqrt(w/(P+w))
  r1 <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E-x1.0))^2)/(2*(P+w)))
  Nt1 <- (r1*N1.0) / (1+alpha.11*N1.0+alpha.12*N2.0)
  r2 <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E-x2.0))^2)/(2*(P+w)))
  Nt2 <- (r2*N2.0) / (1+alpha.22*N2.0+alpha.21*N1.0)
  return(c(Nt1,Nt2,r1,r2))
}
# Simulation of model for t time steps
t <- 40
N <- array(NA,dim=c(t,2))
N <- as.data.frame(N)
colnames(N) <- c("N1","N2")
x <- array(NA,dim=c(t,2))
x <- as.data.frame(x)
colnames(x) <- c("x1","x2")
r <- array(NA,dim=c(t,2))
r <- as.data.frame(r)
colnames(r) <- c("r1","r2")
N$N1[1] <- N1.0
N$N2[1] <- N2.0
x$x1[1] <- x1.0
x$x2[1] <- x2.0
What <- Wmax*sqrt(w/(P+w))
r$r1[1] <- What*(exp((-(((w+(1-h2)*P)/(P+w))*(E.0-x1.0))^2)/(2*(P+w))))
r$r2[1] <- What*(exp((-(((w+(1-h2)*P)/(P+w))*(E.0-x2.0))^2)/(2*(P+w))))
E <- rep(NA, t)
E[1] <- E.0
for (i in 2:t) {
  res <- disc_LV_evol(N1.0=N[i-1,1],N2.0=N[i-1,2],alpha.11=alpha.11,alpha.22=alpha.22,alpha.12=alpha.12,alpha.21=alpha.21,E=E[i-1],x1=x[i-1,1],x2=x[i-1,2],P=P,w=w,Wmax=Wmax,h2=h2)
  N$N1[i] <- res[1]
  N$N2[i] <- res[2]
  r$r1[i] <- res[3]
  r$r2[i] <- res[4]
  # trait change
  d1 <- E[i-1] - x[i-1,1]
  d2 <- E[i-1] - x[i-1,2]
  d1.1 <- k*d1
  d2.1 <- k*d2
  x$x1[i] <- E[i-1] - d1.1
  x$x2[i] <- E[i-1] - d2.1
  # environmental change
  E[i] <- E[i-1] + abs(rnorm(1,0,0.05))
}
```
```{r}
# Plot simulation: ggplot
N$time <- 1:t
dat <- melt(N, id.vars="time")
ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = (1/alpha.11),linetype = "dashed", color = "gray")
```

@@

```{r echo = FALSE}
h2 <- 0
k <- (w+(1-h2)*P)/(P+w)
# Simulation of model for t time steps
Na <- array(NA,dim=c(t,2))
Na <- as.data.frame(Na)
colnames(Na) <- c("N1","N2")
xa <- array(NA,dim=c(t,2))
xa <- as.data.frame(xa)
colnames(xa) <- c("x1","x2")
ra <- array(NA,dim=c(t,2))
ra <- as.data.frame(ra)
colnames(ra) <- c("r1","r2")
Na$N1[1] <- N1.0
Na$N2[1] <- N2.0
xa$x1[1] <- x1.0
xa$x2[1] <- x2.0
What <- Wmax*sqrt(w/(P+w))
ra$r1[1] <- What*(exp((-(((w+(1-h2)*P)/(P+w))*(E.0-x1.0))^2)/(2*(P+w))))
ra$r2[1] <- What*(exp((-(((w+(1-h2)*P)/(P+w))*(E.0-x2.0))^2)/(2*(P+w))))
for (i in 2:t) {
  res <- disc_LV_evol(N1.0=Na[i-1,1],N2.0=Na[i-1,2],alpha.11=alpha.11,alpha.22=alpha.22,alpha.12=alpha.12,alpha.21=alpha.21,E=E[i-1],x1=xa[i-1,1],x2=xa[i-1,2],P=P,w=w,Wmax=Wmax,h2=h2)
  Na$N1[i] <- res[1]
  Na$N2[i] <- res[2]
  ra$r1[i] <- res[3]
  ra$r2[i] <- res[4]
  # trait change
  d1 <- E[i-1] - xa[i-1,1]
  d2 <- E[i-1] - xa[i-1,2]
  d1.1 <- k*d1
  d2.1 <- k*d2
  xa$x1[i] <- E[i-1] - d1.1
  xa$x2[i] <- E[i-1] - d2.1
}
# Plot simulation: ggplot
Na$time <- 1:t
data <- melt(Na, id.vars="time")
ggplot2::ggplot(data, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = (1/alpha.11),linetype = "dashed", color = "gray")
```

@@

```{r}
# Plot simulation: ggplot
dat <- as.data.frame(cbind(log(N$N1),log(N$N2),x$x1,x$x2))
colnames(dat) <- c("N1","N2","x1","x2")
dat$time <- 1:t
df <- data.frame(cbind(dat$N1[2:t],dat$N2[2:t],dat$x1[2:t],dat$x2[2:t]))
colnames(df) <- c("Nt1","Nt2","xt1","xt2")
# Plot
p1 <- ggplot2::ggplot(df, aes(xt1, Nt1)) + geom_point()
p2 <- ggplot2::ggplot(df, aes(xt2, Nt1)) + geom_point()
p3 <- ggplot2::ggplot(df, aes(xt1, Nt2)) + geom_point()
p4 <- ggplot2::ggplot(df, aes(xt2, Nt2)) + geom_point()
p1 + p2 + p3 + p4
```

@@

```{r}
# Plot simulation: ggplot
df$E <- E[2:t]
df$abs_Ex1 <- abs(df$E - df$xt1)
df$abs_Ex2 <- abs(df$E - df$xt2)
# Plot
p1 <- ggplot2::ggplot(df, aes(abs_Ex1, Nt1)) + geom_point()
p2 <- ggplot2::ggplot(df, aes(abs_Ex2, Nt1)) + geom_point()
p3 <- ggplot2::ggplot(df, aes(abs_Ex1, Nt2)) + geom_point()
p4 <- ggplot2::ggplot(df, aes(abs_Ex2, Nt2)) + geom_point()
p1 + p2 + p3 + p4
```

@@

We now onsider the magnitude of trait change from one time step to the next as a driver for species abundances.

```{r}
df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t-1)])
df$dx2 <- abs(dat$x2[2:t] - dat$x2[1:(t-1)])
# Plot
p1 <- ggplot2::ggplot(df, aes(dx1, Nt1)) + geom_point()
p2 <- ggplot2::ggplot(df, aes(dx2, Nt1)) + geom_point()
p3 <- ggplot2::ggplot(df, aes(dx1, Nt2)) + geom_point()
p4 <- ggplot2::ggplot(df, aes(dx2, Nt2)) + geom_point()
p1 + p2 + p3 + p4
```

@@

```{r}
cor(df$dx1, df$abs_Ex1)
cor(df$dx2, df$abs_Ex2)
# Plot
p1 <- ggplot2::ggplot(df, aes(abs_Ex1,dx1)) + geom_point()
p2 <- ggplot2::ggplot(df, aes(abs_Ex2,dx2)) + geom_point()
p1 + p2
```

@@

```{r}
# prepare data in HMSC format
dat <- as.data.frame(cbind(log(N$N1),log(N$N2),x$x1,x$x2))
colnames(dat) <- c("N1","N2","x1","x2")
dat$time <- 1:t
df <- data.frame(cbind(dat$N1[2:t],dat$N2[2:t],dat$x1[2:t],dat$x2[2:t]))
colnames(df) <- c("Nt1","Nt2","xt1","xt2")
df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t-1)])
df$dx2 <- abs(dat$x2[2:t] - dat$x2[1:(t-1)])
# Y matrix
Y <- as.matrix(cbind(df$Nt1,df$Nt2))
# X matrix
XData <- data.frame(cbind(dat$N1[1:(t-1)],dat$N2[1:(t-1)]),E[1:(t-1)],E[1:(t-1)]^2,dat$x1[1:(t-1)],dat$x2[1:(t-1)],abs(dat$x1[2:t] - dat$x1[1:(t-1)]),abs(dat$x2[2:t] - dat$x2[1:(t-1)]),abs(E[1:(t-1)] - dat$x1[1:(t-1)]),abs(E[1:(t-1)] - dat$x2[1:(t-1)]))
colnames(XData) <- c("n1","n2","E","Esq","x1","x2","dx1","dx2","dEx1","dEx2")
# Prepare HMSC model
studyDesign = data.frame(sample = as.factor(1:(t-1)))
rL = HmscRandomLevel(units = studyDesign$sample)
# Design and fit 4 alternative HMSC models
models = list()
for(i in 1:4){
  XFormula = switch(i, ~1, ~E+Esq+x1+x2, ~E+Esq+dEx1+dEx2, ~E+Esq+dx1+dx2)
  m = Hmsc(Y = Y, XData = XData, XFormula = XFormula, studyDesign = studyDesign, ranLevels = list(sample = rL))
  models[[i]] = m
}
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 2000
transient <- 100*thin
verbose <- 500*thin
for(i in 1:4){
  models[[i]] = sampleMcmc(models[[i]],thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
}
```

@@

```{r}
# After Ovaskainen & Abrego 2020, Chapter 7
partition = createPartition(m,nfolds=2,column="sample")
partition.sp = c(1,2)
result = matrix(NA,nrow=3,ncol=4)
for(i in 1:4){
  m = models[[i]]
  # Explanatory power
  preds = computePredictedValues(m)
  MF = evaluateModelFit(hM=m,predY=preds)
  result[1,i] = mean(MF$R2)
  # Predictive power based on cross-validation
  preds = computePredictedValues(m,partition=partition)
  MF = evaluateModelFit(hM=m,predY=preds)
  result[2,i] <- mean(MF$R2)
  # Predictive power based on conditional cross-validation
  preds = computePredictedValues(m,partition=partition,partition.sp=partition.sp,mcmcStep=100)
  MF = evaluateModelFit(hM=m,predY=preds)
  result[3,i] = mean(MF$R2)
}
```

>>>>

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r setup,echo=FALSE,message=FALSE}
library(Hmsc)
library(ggplot2)
library(tidyr)
library(coda)
library(patchwork)
library(jtools)
library(bayesplot)
library(reshape2)
library(ecoevor)
```
```{r,echo=FALSE,message=FALSE,warning=FALSE}
# Wrap long lines
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE,out.width="50%")
```

@@

We simulate population growth under particular parameter values, to generate population dynamics where species interactions, environment, and trait evolution all drive population dynamics. The simulation uses the `disc_LV_evol` command in the `ecoevor` package.

```{r}
set.seed(42)
N0 <- c(10,10)
alpha.11 <- 0.01
alpha.22 <- 0.01
alpha.12 <- 0.005
alpha.21 <- 0.01
alpha <- matrix(c(alpha.11,alpha.21,alpha.12,alpha.22),nrow=2,byrow=FALSE)
E.0 <- 0.8
x.0 <- c(0.1,0.8)
P <- 1
w <- 2
Wmax <- 2
h2 <- 1
k <- (w + (1 - h2) * P)/(P + w)
# Simulation of model for t time steps
t <- 40
N <- array(NA,dim=c(t,length(N0)))
N <- as.data.frame(N)
colnames(N) <- paste0("N",1:length(N0))
N[1,] <- N0
E <- rep(NA, t)
E[1] <- E.0
x <- array(NA,dim=c(t,length(N0)))
x <- as.data.frame(x)
colnames(x) <- paste0("x",1:length(N0))
x[1,] <- x.0
r <- array(NA,dim=c(t,length(N0)))
r <- as.data.frame(r)
colnames(r) <- paste0("r",1:length(N0))
What <- Wmax*sqrt(w/(P+w))
r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
for (i in 2:t) {
  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
  N[i,] <- res$Nt1
  r[i,] <- res$r
  # trait change
  d <- E[i-1] - x[i-1,]
  d1 <- k*d
  x[i,] <- E[i-1] - d1
  # environmental change
  E[i] <- E[i-1] + abs(rnorm(1, 0, 0.05))
}
```
```{r figs1, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1} Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model with competition, a changing environment, and trait evolution. Model parameters are $N_{i,0}$ = 10, $\\alpha_{ii}$ = 0.01, $\\alpha_{12}$ = 0.005, $\\alpha_{21}$ = 0.01, and $h^2 = 1$."}
# Plot simulation: ggplot
N$time <- 1:t
dat <- melt(N, id.vars="time")
ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + ylab("N") + theme_classic()
```

@@

```{r echo = FALSE}
h2 <- 0
k <- (w+(1-h2)*P)/(P+w)
# Simulation of model for t time steps
Na <- array(NA,dim=c(t,length(N0)))
Na <- as.data.frame(Na)
colnames(Na) <- paste0("N",1:length(N0))
xa <- array(NA,dim=c(t,length(N0)))
xa <- as.data.frame(xa)
colnames(xa) <- paste0("x",1:length(N0))
ra <- array(NA,dim=c(t,length(N0)))
ra <- as.data.frame(ra)
colnames(ra) <- paste0("r",1:length(N0))
Na[1,] <- N0
xa[1,] <- x.0
What <- Wmax*sqrt(w/(P+w))
ra[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-xa[1,]))^2)/(2*(P+w)))
for (i in 2:t) {
  res <- disc_LV_evol(N0=Na[i-1,],alpha=alpha,E=E[i-1],x=xa[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
  Na[i,] <- res$Nt1
  ra[i,] <- res$r
  # trait change
  d <- E[i-1] - x[i-1,]
  d1 <- k*d
  xa[i,] <- E[i-1] - d1
  # environmental change
  E[i] <- E[i-1] + abs(rnorm(1, 0, 0.05))
}
```
```{r figs2, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs2} Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model with competition, a changing environment, and no trait evolution. Model parameters are $N_{i,0}$ = 10, $\\alpha_{ii}$ = 0.01, $\\alpha_{12}$ = 0.005, $\\alpha_{21}$ = 0.01, and $h^2 = 0$."}
# Plot simulation: ggplot
Na$time <- 1:t
data <- melt(Na, id.vars="time")
ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + ylab("N") + theme_classic()
```

@@

```{r figs3, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs3} Population size (logarithm) in species 1 and 2 at one time step $N_{i,t+1}$ as a function of trait value of Species 1 and Species 2 in the previous time step $x_t$."}
# Plot simulation: ggplot
dat <- as.data.frame(cbind(log(N$N1),log(N$N2),x$x1,x$x2))
colnames(dat) <- c("N1","N2","x1","x2")
dat$time <- 1:t
df <- data.frame(cbind(dat$N1[2:t],dat$N2[2:t],dat$x1[2:t],dat$x2[2:t]))
colnames(df) <- c("Nt1","Nt2","xt1","xt2")
# Plot
p1 <- ggplot2::ggplot(df, aes(xt1, Nt1)) + geom_point() + theme_classic()
p2 <- ggplot2::ggplot(df, aes(xt2, Nt1)) + geom_point() + theme_classic()
p3 <- ggplot2::ggplot(df, aes(xt1, Nt2)) + geom_point() + theme_classic()
p4 <- ggplot2::ggplot(df, aes(xt2, Nt2)) + geom_point() + theme_classic()
p1 + p2 + p3 + p4
```

@@

```{r figs4, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs43} Population size (logarithm) in species 1 and 2 at one time step $N_{i,t+1}$ as a function of the absolute value of distance between the environmental optimum trait value given by $E_t$ and the trait value of Species 1 and Species 2 $x_t$."}
# Plot simulation: ggplot
df$E <- E[2:t]
df$abs_Ex1 <- abs(df$E - df$xt1)
df$abs_Ex2 <- abs(df$E - df$xt2)
# Plot
p1 <- ggplot2::ggplot(df, aes(abs_Ex1, Nt1)) + geom_point() + theme_classic()
p2 <- ggplot2::ggplot(df, aes(abs_Ex2, Nt1)) + geom_point() + theme_classic()
p3 <- ggplot2::ggplot(df, aes(abs_Ex1, Nt2)) + geom_point() + theme_classic()
p4 <- ggplot2::ggplot(df, aes(abs_Ex2, Nt2)) + geom_point() + theme_classic()
p1 + p2 + p3 + p4
```

@@

We now consider the magnitude of trait change from one time step to the next as a driver for species abundances.

```{r figs5, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs5} Population size (logarithm) in species 1 and 2 at one time step $N_{i,t+1}$ as a function of change in trait value of Species 1 and Species 2 in the previous time step $|\\Delta x_{t,t-1}|$."}
df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t-1)])
df$dx2 <- abs(dat$x2[2:t] - dat$x2[1:(t-1)])
# Plot
p1 <- ggplot2::ggplot(df, aes(dx1, Nt1)) + geom_point() + theme_classic()
p2 <- ggplot2::ggplot(df, aes(dx2, Nt1)) + geom_point() + theme_classic()
p3 <- ggplot2::ggplot(df, aes(dx1, Nt2)) + geom_point() + theme_classic()
p4 <- ggplot2::ggplot(df, aes(dx2, Nt2)) + geom_point() + theme_classic()
p1 + p2 + p3 + p4
```

@@

```{r}
cor(df$dx1, df$abs_Ex1)
cor(df$dx2, df$abs_Ex2)
```

```{r figs6, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs6} Relationship between change in trait value of Species 1 and Species 2 in the previous time step $|\\Delta x_{t,t-1}|$ and distance from environmental optimum trait value to species trait value $|E_t - x_t|$."}
# Plot
p1 <- ggplot2::ggplot(df, aes(abs_Ex1,dx1)) + geom_point() + theme_classic()
p2 <- ggplot2::ggplot(df, aes(abs_Ex2,dx2)) + geom_point() + theme_classic()
p1 + p2
```

@@

```{r results='hide'}
# prepare data in HMSC format
dat <- as.data.frame(cbind(log(N$N1),log(N$N2),x$x1,x$x2))
colnames(dat) <- c("N1","N2","x1","x2")
dat$time <- 1:t
df <- data.frame(cbind(dat$N1[2:t],dat$N2[2:t],dat$x1[2:t],dat$x2[2:t]))
colnames(df) <- c("Nt1","Nt2","xt1","xt2")
df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t-1)])
df$dx2 <- abs(dat$x2[2:t] - dat$x2[1:(t-1)])
# Y matrix
Y <- as.matrix(cbind(df$Nt1,df$Nt2))
# X matrix
XData <- data.frame(cbind(dat$N1[1:(t-1)],dat$N2[1:(t-1)]),E[1:(t-1)],E[1:(t-1)]^2,dat$x1[1:(t-1)],dat$x2[1:(t-1)],abs(dat$x1[2:t] - dat$x1[1:(t-1)]),abs(dat$x2[2:t] - dat$x2[1:(t-1)]),abs(E[1:(t-1)] - dat$x1[1:(t-1)]),abs(E[1:(t-1)] - dat$x2[1:(t-1)]))
colnames(XData) <- c("n1","n2","E","Esq","x1","x2","dx1","dx2","dEx1","dEx2")
# Prepare HMSC model
studyDesign = data.frame(sample = as.factor(1:(t-1)))
rL = HmscRandomLevel(units = studyDesign$sample)
# Design and fit 4 alternative HMSC models
models = list()
for(i in 1:4){
  XFormula = switch(i, ~1, ~E+Esq+x1+x2, ~E+Esq+dEx1+dEx2, ~E+Esq+dx1+dx2)
  m = Hmsc(Y = Y, XData = XData, XFormula = XFormula, studyDesign = studyDesign, ranLevels = list(sample = rL))
  models[[i]] = m
}
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 2000
transient <- 100*thin
verbose <- 500*thin
for(i in 1:4){
  models[[i]] = sampleMcmc(models[[i]],thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
}
```

@@

```{r results='hide'}
# After Ovaskainen & Abrego 2020, Chapter 7
partition = createPartition(m,nfolds=2,column="sample")
partition.sp = c(1,2)
result = matrix(NA,nrow=3,ncol=4)
for(i in 1:4){
  m = models[[i]]
  # Explanatory power
  preds = computePredictedValues(m)
  MF = evaluateModelFit(hM=m,predY=preds)
  result[1,i] = mean(MF$R2)
  # Predictive power based on cross-validation
  preds = computePredictedValues(m,partition=partition)
  MF = evaluateModelFit(hM=m,predY=preds)
  result[2,i] <- mean(MF$R2)
  # Predictive power based on conditional cross-validation
  preds = computePredictedValues(m,partition=partition,partition.sp=partition.sp,mcmcStep=100)
  MF = evaluateModelFit(hM=m,predY=preds)
  result[3,i] = mean(MF$R2)
}
```
